 EDA_ Manuscript

### 2025-04-05 code/analysis-code/Statistical_Analysis.qmd
Project update

*Boxplot of Birthweight by IPTp Treatment Arm*

```{r}
ggplot(promo_data_clean, aes(x = study_arm, y = birth_weight, fill = study_arm)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Birthweight by IPTp Treatment Arm",
       x = "Treatment Arm",
       y = "Birthweight (kg)") +
  theme_minimal() +
  theme(legend.position = "none")

@@

```{r}
# Save the cleaned data (if not already saved)
write.csv(promo_data_clean, here::here("data", "clean", "PROMO_Data_clean.csv"), row.names = FALSE)

# Read in the cleaned data
promo_data_clean <- read.csv(here::here("data", "clean", "PROMO_Data_clean.csv"))

# Load necessary libraries
library(dplyr)
library(ggplot2)
library(broom)

### Data Preparation

# 1. Create a composite adverse birth outcome variable
#    Adverse outcome is defined as: 
#    - Preterm birth: preterm_births_count > 0 
#    OR stillbirth: stillbirth_bin == 1 
#    OR low birth weight: birth_weight < 2.5 kg
promo_data_clean <- promo_data_clean %>%
  mutate(low_birth_weight = ifelse(birth_weight < 2.5, 1, 0),
         adverse_birth_outcome = ifelse(preterm_births_count > 0 | stillbirth_bin == 1 | low_birth_weight == 1, 1, 0))

# 2. Create a subgroup variable for maternal age (<25 years vs. >=25 years)
promo_data_clean <- promo_data_clean %>%
  mutate(age_group = ifelse(age_at_enrollment_years < 25, "Young", "Older"))

### Research Question 1: 
### "Does the type of IPTp regimen modify the association between malaria episode frequency and adverse birth outcomes in Ugandan pregnant women?"

# Fit a logistic regression model with an interaction between total malaria episodes and study_arm
model_interaction <- glm(adverse_birth_outcome ~ total_malaria_episodes * study_arm + 
                           age_at_enrollment_years + gravidity + education_level,
                         family = binomial(link = "logit"),
                         data = promo_data_clean)

# Summarize and tidy the model output (exponentiating coefficients to yield odds ratios)
summary(model_interaction)
tidy_model_interaction <- tidy(model_interaction, exponentiate = TRUE, conf.int = TRUE)
print(tidy_model_interaction)

### Research Question 2: 
### "Among younger Ugandan pregnant women, is increased gravidity associated with a reduced risk of adverse birth outcomes?"

# Subset the data for young mothers (age < 25)
young_mothers <- promo_data_clean %>% filter(age_group == "Young")

# Fit a logistic regression model within the young mothers subgroup 
# to examine the association between gravidity and adverse birth outcomes
model_gravidity_young <- glm(adverse_birth_outcome ~ gravidity + total_malaria_episodes + education_level,
                             family = binomial(link = "logit"),
                             data = young_mothers)

# Summarize and tidy the subgroup model output
summary(model_gravidity_young)
tidy_model_gravidity_young <- tidy(model_gravidity_young, exponentiate = TRUE, conf.int = TRUE)
print(tidy_model_gravidity_young)

### Additional Exploratory Plots (Optional)

# Plot the distribution of total malaria episodes by study arm
ggplot(promo_data_clean, aes(x = study_arm, y = total_malaria_episodes)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Malaria Episode Frequency by Treatment Arm", 
       x = "Treatment Arm", y = "Total Malaria Episodes")

# Plot the proportion of adverse outcomes by gravidity among young mothers
ggplot(young_mothers, aes(x = gravidity, fill = factor(adverse_birth_outcome))) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Adverse Outcomes by Gravidity (Young Mothers)",
       x = "Gravidity", y = "Proportion", fill = "Adverse Outcome")

@@

@@

*Predicted Probability of Adverse Outcome by Gravidity (Age \< 25)*

@@

# Create a scatter/line plot of predicted probabilities by gravidity
ggplot(predicted_probs, aes(x = x, y = predicted)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, fill = "blue") +
  labs(title = "Predicted Probability of Adverse Outcome by Gravidity (Age < 25)",
       x = "Gravidity",
       y = "Predicted Probability") +
  theme_minimal()

>>>>

@@

@@

*Table 6: Adjusted Odds Ratios for Adverse Birth Outcomes Among Young Pregnant Women (<25 Years)*

```{r}
# Load required libraries
library(broom)
library(dplyr)
library(knitr)
library(kableExtra)

# Assume model_gravidity is your fitted logistic regression model for women < 25
# For example:
# model_gravidity <- glm(adverse_birth_outcome ~ gravidity + total_malaria_episodes + 
#                          study_arm + education_level,
#                        family = binomial(link = "logit"),
#                        data = promo_data_young)

# Tidy the model output, exponentiating coefficients to get odds ratios
tidy_model_gravidity <- tidy(model_gravidity, exponentiate = TRUE, conf.int = TRUE)

# Create a summary table without the intercept and with neat variable names
result_table <- tidy_model_gravidity %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = case_when(
      term == "gravidity" ~ "Gravidity",
      term == "total_malaria_episodes" ~ "Total Malaria Episodes",
      term == "study_armSP" ~ "Treatment Arm (SP)",
      term == "education_levelSecondary" ~ "Secondary Education",
      term == "education_levelTertiary" ~ "Tertiary Education",
      term == "education_levelUniversity" ~ "University Education",
      TRUE ~ term
    ),
    `Odds Ratio (95% CI)` = paste0(round(estimate, 2),
                                   " (", round(conf.low, 2), ", ", round(conf.high, 2), ")")
  ) %>%
  select(term, `Odds Ratio (95% CI)`, p.value) %>%
  rename(Variable = term, `p-value` = p.value)

# Print the table using kable
result_table %>%
  kable(format = "markdown", caption = "")

@@

```{r}
# Tidy the model output and exponentiate coefficients to get odds ratios
tidy_model_gravidity <- tidy(model_gravidity, exponentiate = TRUE, conf.int = TRUE)

# Create a summary table, remove the intercept, and recode variable names
result_table <- tidy_model_gravidity %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = dplyr::recode(term,
                         "gravidity" = "Gravidity",
                         "total_malaria_episodes" = "Total Malaria Episodes",
                         "study_armSP" = "Treatment Arm (SP)",
                         "education_levelSecondary" = "Secondary Education",
                         "education_levelTertiary" = "Tertiary Education",
                         "education_levelUniversity" = "University Education"),
    `Odds Ratio (95% CI)` = paste0(round(estimate, 2),
                                   " (", round(conf.low, 2), ", ", round(conf.high, 2), ")")
  ) %>%
  select(term, `Odds Ratio (95% CI)`, p.value) %>%
  rename(Variable = term, `p-value` = p.value)

# Print the table using kable with a descriptive caption
result_table %>%
  kable(format = "markdown", 
        caption = "Adjusted Odds Ratios for Adverse Birth Outcomes Among Young Pregnant Women (<25 Years)")

*Predicted Probability of Adverse Outcome by Gravidity (Age < 25)*

@@

# Create a scatter/line plot of predicted probabilities by gravidity
ggplot(predicted_probs, aes(x = x, y = predicted)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, fill = "blue") +
  labs(title = "",
       x = "Gravidity",
       y = "Predicted Probability") +
  theme_minimal()

### 2025-04-05 code/eda-code/eda_manuscript.qmd
Project update

@@

*Correlation Matrix (Numeric Variables)*
```{r}
# Correlation plot
numeric_vars <- promo_data_clean %>% 
  select(where(is.numeric))

cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")

corrplot::corrplot(cor_matrix, method = "circle", type = "lower")

*Group-wise Summary*
```{r}
# Grouped summary by treatment group (replace 'treatment_group' with actual variable)
promo_data_clean %>% 
  group_by(treatment_group) %>% 
  summarise(across(where(is.numeric), list(mean = mean, sd = sd), na.rm = TRUE))

>>>>

*Boxplot of Birthweight by IPTp Treatment Arm*

```{r}
ggplot(promo_data_clean, aes(x = study_arm, y = birth_weight, fill = study_arm)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Birthweight by IPTp Treatment Arm",
       x = "Treatment Arm",
       y = "Birthweight (kg)") +
  theme_minimal() +
  theme(legend.position = "none")

```{r}
# Save the cleaned data (if not already saved)
write.csv(promo_data_clean, here::here("data", "clean", "PROMO_Data_clean.csv"), row.names = FALSE)

# Read in the cleaned data
promo_data_clean <- read.csv(here::here("data", "clean", "PROMO_Data_clean.csv"))

# Load necessary libraries
library(dplyr)
library(ggplot2)
library(broom)

### Data Preparation

# 1. Create a composite adverse birth outcome variable
#    Adverse outcome is defined as: 
#    - Preterm birth: preterm_births_count > 0 
#    OR stillbirth: stillbirth_bin == 1 
#    OR low birth weight: birth_weight < 2.5 kg
promo_data_clean <- promo_data_clean %>%
  mutate(low_birth_weight = ifelse(birth_weight < 2.5, 1, 0),
         adverse_birth_outcome = ifelse(preterm_births_count > 0 | stillbirth_bin == 1 | low_birth_weight == 1, 1, 0))

# 2. Create a subgroup variable for maternal age (<25 years vs. >=25 years)
promo_data_clean <- promo_data_clean %>%
  mutate(age_group = ifelse(age_at_enrollment_years < 25, "Young", "Older"))

### Research Question 1: 
### "Does the type of IPTp regimen modify the association between malaria episode frequency and adverse birth outcomes in Ugandan pregnant women?"

# Fit a logistic regression model with an interaction between total malaria episodes and study_arm
model_interaction <- glm(adverse_birth_outcome ~ total_malaria_episodes * study_arm + 
                           age_at_enrollment_years + gravidity + education_level,
                         family = binomial(link = "logit"),
                         data = promo_data_clean)

# Summarize and tidy the model output (exponentiating coefficients to yield odds ratios)
summary(model_interaction)
tidy_model_interaction <- tidy(model_interaction, exponentiate = TRUE, conf.int = TRUE)
print(tidy_model_interaction)

### Research Question 2: 
### "Among younger Ugandan pregnant women, is increased gravidity associated with a reduced risk of adverse birth outcomes?"

# Subset the data for young mothers (age < 25)
young_mothers <- promo_data_clean %>% filter(age_group == "Young")

# Fit a logistic regression model within the young mothers subgroup 
# to examine the association between gravidity and adverse birth outcomes
model_gravidity_young <- glm(adverse_birth_outcome ~ gravidity + total_malaria_episodes + education_level,
                             family = binomial(link = "logit"),
                             data = young_mothers)

# Summarize and tidy the subgroup model output
summary(model_gravidity_young)
tidy_model_gravidity_young <- tidy(model_gravidity_young, exponentiate = TRUE, conf.int = TRUE)
print(tidy_model_gravidity_young)

### Additional Exploratory Plots (Optional)

# Plot the distribution of total malaria episodes by study arm
ggplot(promo_data_clean, aes(x = study_arm, y = total_malaria_episodes)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Malaria Episode Frequency by Treatment Arm", 
       x = "Treatment Arm", y = "Total Malaria Episodes")

# Plot the proportion of adverse outcomes by gravidity among young mothers
ggplot(young_mothers, aes(x = gravidity, fill = factor(adverse_birth_outcome))) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Adverse Outcomes by Gravidity (Young Mothers)",
       x = "Gravidity", y = "Proportion", fill = "Adverse Outcome")

@@
